---
title: "Basic Data Visualization in R"
title-short: "R DataViz"
author: 
   - "Siddhartha Bora, Assistant Professor, West Virginia University"
date: "June 26, 2024"
date-format: long
format: 
  revealjs:
    auto-strech: true
    code-link: true
    incremental: false
    embed-resources: true
    grow: true
    smaller: true
    scrollable: true
---



## Outline

- Some thoughts on data wrangling
- An introduction to the workflow of data visualization in R.
- Visualization examples.
- A basic introduction to Quarto


## Disclaimers

-   The content of this presentation is based on the author's experience and knowledge of R, and represent an opinionated view of data visualizations.
-   R is vast and the content of this presentation is not exhaustive of all the data visualization features available in R. The author recognizes that there are many other ways to visualize data in R.
-   The contents of this presentation are tailored to create static data visualizations for the competition.
-   The views and opinions expressed in this presentation are those of the author and do not necessarily reflect the official policy or position of any other agency, organization, employer or company.




## Tidy Data Principles

1.  Each variable is a column; each column is a variable.
2.  Each observation is a row; each row is an observation.
3.  Each value is a cell; each cell is a single value.

Source: [R for Data Science (2e), Hadley Wickham, Mine Cetinkaya-Rundel, Garret Golemund](https://r4ds.hadley.nz/data-tidy)

## Long vs Wide Data

-   Long form
```{r}
library(tidyverse)
# long data small example
df_long<-tibble(
  student_id=c(1,1, 1, 2, 2,2),
  subject=c("english","math","science", "english","math","science"),
  test_score=c(70, 87, 100,  78, 65, 83)
)
df_long
```

-   Wide form

```{r}
# convert long data to wide data
df_wide<-df_long |> 
  pivot_wider(names_from = subject, values_from = test_score)

df_wide
```
-   Useful functions: `pivot_longer` and `pivot_wider` in the `tidyverse` 
- long data is typically useful for visualization.

## Data Preprocessing takes most of the time

-   Data wrangling is the process of cleaning and transforming raw data into a format that is more suitable for analysis and visualization.
-   Data wrangling is often the most time-consuming part of the data analysis process.
-   The `tidyverse` package in R provides a set of tools for data wrangling that are designed to work together.
-  I am using a fraction of the data from the Local Food Economics Warehouse.


## Data Wrangling 

::: panel-tabset
#### Code

```{r}
#| echo: true
#| eval: false
#| warning: false

library("tidyverse")
library("kableExtra")

# load the data
df_foodaccess<-read_csv("./data/df_foodaccess.csv")
meta_foodaccess<-read_csv("./data/meta_foodaccess.csv")

# let's keep only overall_food_insecurity_rate and child_food_insecurity_rate
mydata<-df_foodaccess |>
  filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate"))

# write data into a csv
write_csv(mydata, "./data/food_insecurity_dataviz.csv")

```

#### Metadata

```{r}
#| label: cleaned-dataset
#| echo: false
#| eval: true
#| warning: false


library("tidyverse")
library("kableExtra")

# load the data
df_foodaccess<-read_csv("./data/df_foodaccess.csv")
meta_foodaccess<-read_csv("./data/meta_foodaccess.csv")

# let's keep only overall_food_insecurity_rate and child_food_insecurity_rate
df_foodsecurity<-df_foodaccess |>
  filter(variable_name %in% c("overall_food_insecurity_rate", 
                              "child_food_insecurity_rate")
         )
meta_foodsecurity<-meta_foodaccess |>
  filter(variable_name %in% c("overall_food_insecurity_rate", 
                              "child_food_insecurity_rate")
         )

# write data into a csv
write_csv(df_foodsecurity, "./data/food_insecurity_dataviz.csv")

# metadata
metadata<-meta_foodaccess |>
  filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate"))|>
  select(user_friendly_variable_name,
         variable_definition,
         source,
         citation) |>
  kable()|>
  kable_styling("striped", full_width = F)

metadata

```
:::

## Layered Grammar of Graphics

"All plots are composed of the data, the information you want to visualise, and a mapping, the description of how the data’s variables are mapped to aesthetic attributes."

-   Mapping components: layers (geoms and stats), scales, coords, facets,and themes.


Source: [ggplot2: Elegant Graphics for Data Analysis (3e), Hadley Wickham, Mine Cetinkaya-Rundel, Garret Golemund] (https://ggplot2-book.org/introduction)

## Example: Food Insecurity in West Virginia

::: panel-tabset
#### Code

```{r}
#| label: wv-dataprep-chunk1
#| echo: true
#| eval: false
#| warning: false
# Subset data for WV
df_wv <- df_foodsecurity |> 
  filter(fips == "54") |>
  mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                     "child_food_insecurity_rate"),
         value = 100*value)

# examine data
head(df_wv)

```

I try to subset data for West Virginia.

#### Output

```{r}
#| label: wv-dataprep-chunk2
#| echo: false
#| eval: true
#| warning: false
# Subset data for WV
# Subset data for WV
df_wv <- df_foodsecurity |> 
  filter(fips == "54") |>
  mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                     "child_food_insecurity_rate"),
         value = 100*value)

head(df_wv)

```
:::

## Example: Food Insecurity in WV


I try to draw a basic lineplot.
```{r}
#| label: basic-lineplot-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3


# line plot
df_wv |>
  filter(variable_name == "overall_food_insecurity_rate") |>
  ggplot(aes(x = year, y = value)) +  
  geom_line()


```


## Example: Food Insecurity in WV

Let's try plotting both variables

```{r}
#| label: wv-lineplot2-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3

# create line plot for two variables
df_wv |>
  ggplot(aes(x = year, y = value, color = variable_name)) +  
  geom_line()

```


## Example: Food Insecurity in WV

Let's try doing barplots.

```{r}
#| label: wv-barplot1-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3
# create barchart for two variables 
df_wv |>
  ggplot(aes(x = year, y = value, fill = variable_name)) +  
  geom_col(position = "dodge")


```


## Improving the Plot

-  I decided to make a barplot.
- The plot works, but needs some improvement.
- Improvements can take many forms, such as:
    - Adding labels
    - Adding title
    - Adding subtitle
    - Adding caption
    - Changing the color scheme
    - Changing the theme
    
- I will try to improve it layer by layer

## Treat the Plot as an Object



```{r}
#| label: object-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3

# treat the plot as an object
plot_1 <- df_wv |>
  filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate")) |>
  ggplot(aes(x = year, y = value, fill = variable_name)) +  
  geom_col(position = "dodge")

plot_2<- plot_1+
  labs(title="Food Insecurity Rate in West Virginia", x="", y="%")

plot_2

```

## Further Improvements


```{r}
#| label: labels-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3

plot_3 <- plot_2+
  scale_x_continuous(breaks = seq(2011, 2021, 1))+
  geom_text(aes(label = paste0(round(value, 1), "%")), 
            position = position_dodge(width = 1), 
            vjust = -0.5, size = 3)+
  scale_fill_discrete(name = "", labels = c("Overall", "Child"))

plot_3
```


## Improve theme


```{r}
#| label: improvetheme-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 3

plot_4 <- plot_3+
  theme_minimal()+
  theme(legend.position = "bottom",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

plot_4

```



## Faceting 


::: panel-tabset
#### Code

```{r}
#| label: facet-chunk1
#| echo: true
#| eval: false
#| warning: false

# let's make graphs for both OH and WV
df_oh_wv <- df_foodsecurity |> 
  filter(fips %in% c("39","54")) |>
  mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                     "child_food_insecurity_rate"),
         value = 100*value)


plot_5 <- df_oh_wv |>
  filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate")) |>
  ggplot(aes(x = year, y = value, fill = variable_name)) +  
  geom_col(position = "dodge")+
  labs(title="Food Insecurity Rates in the Ohio and West Virginia", x="", y="%")+
  # break each year
  scale_x_continuous(breaks = seq(2011, 2021, 1))+
  # add value labels
  geom_text(aes(label = paste0(round(value, 1), "%")), 
            position = position_dodge(width = 1), 
            vjust = -0.5, size = 3)+
  # legend labels and name
  scale_fill_discrete(name = "", labels = c("Overall", "Child"))+
  theme_minimal()+
  theme(legend.position = "bottom",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  facet_wrap(~state_name)

plot_5


```

#### Output

```{r}
#| label: facet-chunk2
#| echo: false
#| eval: true
#| warning: false

# let's make graphs for both OH and WV
df_oh_wv <- df_foodsecurity |> 
  filter(fips %in% c("39","54")) |>
  # variable name factor relevel
  mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                     "child_food_insecurity_rate"),
         value = 100*value)

# treat the plot as an object
plot_3 <- df_oh_wv |>
  filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate")) |>
  ggplot(aes(x = year, y = value, fill = variable_name)) +  
  geom_col(position = "dodge")+
  labs(title="Food Insecurity Rates in the Ohio and West Virginia", x="", y="%")+
  # break each year
  scale_x_continuous(breaks = seq(2011, 2021, 1))+
  # add value labels
  geom_text(aes(label = paste0(round(value, 1), "%")), 
            position = position_dodge(width = 1), 
            vjust = -0.5, size = 3)+
  # legend labels and name
  scale_fill_discrete(name = "", labels = c("Overall", "Child"))+
  theme_minimal()+
  theme(legend.position = "bottom",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust = 0.5),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  facet_wrap(~state_name)

plot_3
```
:::


## Reusing Code as functions

::: panel-tabset
#### Code

```{r}
#| label: barplot-functions-chunk1
#| echo: true
#| eval: false
#| warning: false

# Make a function to create the plot for any state
plot_food_insecurity_state <- function(state_fips){
  df_state <- df_foodsecurity |> 
    filter(fips == state_fips) |>
    # variable name factor relevel
    mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                       "child_food_insecurity_rate"))
  
  state_name <- df_state |>
    pull(state_name) |>
    unique()
  
  # treat the plot as an object
  plot_state <- df_state |>
    filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate")) |>
    ggplot(aes(x = year, y = 100*value, fill = variable_name)) +  
    geom_col(position = "dodge")+
    labs(title=paste("Food Insecurity Rates in", state_name), x="", y="%")+
    # break each year
    scale_x_continuous(breaks = seq(2011, 2021, 1))+
    ylim(0, 40)+
    # add value labels
    geom_text(aes(label = paste0(round(100*value, 1), "%")), 
              position = position_dodge(width = 1), 
              vjust = -0.5, size = 3)+
    # legend labels and name
    scale_fill_discrete(name = "", labels = c("Overall", "Child"))+
    theme_minimal()+
    theme(legend.position = "bottom",
          legend.title = element_text(size = 12),
          legend.text = element_text(size = 10),
          plot.title = element_text(size = 14, hjust = 0.5),
          axis.title = element_text(size = 12),
          axis.text = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  
  return(plot_state)
}
  

# Let's try for Virginia
plot_food_insecurity_state("51")


```

#### Output

```{r}
#| label: barplot-functions-chunk2
#| echo: false
#| eval: true
#| warning: false

# Make a function to create the plot for any state
plot_food_insecurity_state <- function(state_fips){
  df_state <- df_foodsecurity |> 
    filter(fips == state_fips) |>
    # variable name factor relevel
    mutate(variable_name = fct_relevel(variable_name, "overall_food_insecurity_rate", 
                                       "child_food_insecurity_rate"))
  
  state_name <- df_state |>
    pull(state_name) |>
    unique()
  
  # treat the plot as an object
  plot_state <- df_state |>
    filter(variable_name %in% c("overall_food_insecurity_rate", "child_food_insecurity_rate")) |>
    ggplot(aes(x = year, y = 100*value, fill = variable_name)) +  
    geom_col(position = "dodge")+
    labs(title=paste("Food Insecurity Rates in", state_name), x="", y="%")+
    # break each year
    scale_x_continuous(breaks = seq(2011, 2021, 1))+
    ylim(0, 40)+
    # add value labels
    geom_text(aes(label = paste0(round(100*value, 1), "%")), 
              position = position_dodge(width = 1), 
              vjust = -0.5, size = 3)+
    # legend labels and name
    scale_fill_discrete(name = "", labels = c("Overall", "Child"))+
    theme_minimal()+
    theme(legend.position = "bottom",
          legend.title = element_text(size = 12),
          legend.text = element_text(size = 10),
          plot.title = element_text(size = 14, hjust = 0.5),
          axis.title = element_text(size = 12),
          axis.text = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
  
  return(plot_state)
}
  

# Let's try for Virginia
plot_food_insecurity_state("51")
```
:::


## Arranging Plots


```{r}
#| label: arrange-chunk1
#| echo: true
#| eval: true
#| warning: false
#| output-location: fragment
#| tidy: true
#| tidy.opts: { width.cutoff: 60 }
#| fig-height: 5
# Arranging plots in a grid
library("gridExtra")
plot_wv<-plot_food_insecurity_state("54")
plot_va<-plot_food_insecurity_state("51")

# grid arrange and merge legends
grid.arrange(plot_wv, plot_va, ncol=2)

```


## Map Example 1: State Level Food Insecurity Rate

Prepare data for plotting the map.

```{r}
#| output: false
#| echo: true


library("tidyverse")
library("sf")
library("tigris")
library("kableExtra")

df_foodinsecurity<-read_csv("./data/food_insecurity_dataviz.csv")

# let's plot a map
df_states<-df_foodinsecurity |>
  filter(as.numeric(fips)<=56 & year ==2021 & state_name!="US")|>
  filter(variable_name == "overall_food_insecurity_rate")


states<-tigris::states(filter_by = c(xmin=-125, xmax=-66, ymin=24, ymax=50))|>
  left_join(df_states, by = join_by(GEOID==fips))

```

## Map of Overall Food Security in the US

::: panel-tabset
#### Code

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-cap: "US Food Insecurity Rate"

# plot the map
ggplot(states, aes(fill=value))+
  geom_sf()+
  scale_fill_viridis_c()+
  labs(title="Food Insecurity Rate in the US", fill="Rate")+
  theme_minimal()

```

#### Output

```{r}
#| label: fig-us-states-food-insecurity
#| echo: false
#| eval: true
#| warning: false
#| fig-cap: "Overall Food Insecurity Rate, US States, 2021"

# plot the map
ggplot(states, aes(fill=value))+
  geom_sf()+
  scale_fill_viridis_c()+
  labs(title="Food Insecurity Rate in the US", fill="Rate")+
  theme_minimal()

```
:::

## County Level Map of West Virginia

::: panel-tabset
#### Code

```{r}
#| label: wvmap-chunk1
#| echo: true
#| eval: false
#| warning: false

# county level maps
df_states<-mydata |>
  filter(as.numeric(fips)>1000 & year ==2021 & state_name!="US")|>
  filter(variable_name == "overall_food_insecurity_rate")
wv_counties <- counties(state="WV")|>
  left_join(df_states, by = join_by(GEOID==fips))


# plot the map
ggplot(wv_counties, aes(fill=value))+
  geom_sf()+
  scale_fill_viridis_c()+
  labs(title="Food Insecurity Rate in WV", fill="Rate")+
  theme_minimal()



```
```{r}
#| output: false
#| echo: false
# county level maps
df_states<-df_foodinsecurity |>
  filter(as.numeric(fips)>1000 & year ==2021 & state_name!="US")|>
  filter(variable_name == "overall_food_insecurity_rate")
wv_counties <- counties(state="WV")|>
  left_join(df_states, by = join_by(GEOID==fips))

```

#### Output

```{r}
#| label: wvmap-chunk2
#| echo: false
#| eval: true
#| warning: false

# plot the map
ggplot(wv_counties, aes(fill=value))+
  geom_sf()+
  scale_fill_viridis_c()+
  labs(title="County Level Food Insecurity Rate in WV, 2021", fill="Rate")+
  theme_minimal()


```
:::


## Generalize using a function and make Ohio map

::: panel-tabset
#### Code

```{r}
#| label: ohiomap-chunk1
#| echo: true
#| eval: false
#| warning: false

plot_state_map<-function(df, state_name, variable_name, year){
  df_states<-df |>
    filter(as.numeric(fips)>1000 & year ==year & state_name==state_name)|>
    filter(variable_name == variable_name)
  
  counties <- counties(state=state_name)|>
    left_join(df_states, by = join_by(GEOID==fips))
  
  # plot the map
  ggplot(counties, aes(fill=value))+
    geom_sf()+
    scale_fill_viridis_c()+
    labs(title=paste(variable_name, "in", state_name), fill="Rate")+
    theme_minimal()
}

plot_state_map(df_foodinsecurity, "Ohio", "child_food_insecurity_rate", 2021)

```

#### Output

```{r}
#| label: ohiomap-chunk2
#| echo: false
#| eval: true
#| warning: false
plot_state_map<-function(df, state_name, variable_name, year){
  df_states<-df |>
    filter(as.numeric(fips)>1000 & year ==year & state_name==state_name)|>
    filter(variable_name == variable_name)
  
  counties <- counties(state=state_name)|>
    left_join(df_states, by = join_by(GEOID==fips))
  
  # plot the map
  ggplot(counties, aes(fill=value))+
    geom_sf()+
    scale_fill_viridis_c()+
    labs(title=paste(variable_name, "in", state_name), fill="Rate")+
    theme_minimal()
}

plot_state_map(df_foodinsecurity, "Ohio", "child_food_insecurity_rate", 2021)

```
:::




## Infographics using Quarto

-  Quarto has been around a while.
-  Powerful tool for creating dynamic documents.
-  Can be useful for static visualizations as well.
-  In particular, visualization objects created in R can be easily embedded in Quarto documents.
-  Put together everything for infographic.

## Parameterized Infographics/Reports

- A powerful feature of Quarto is the ability to parameterize your documents. 
- This allows you to create a single document that can be rendered multiple times with different inputs. 
- While this competition is looking for static visualizations, parameterized infographics can still add dynamism to your submission while keeping the end product static.
- Let's see an example of how you can create a parameterized infographic in Quarto.




## Additional Resources: R Workflow Related Resources

-   What They Forgot to Teach You About R: The stuff you need to know about R, besides data analysis. by Jennifer Bryan, Jennifer Bryan, Jim Hester, Shannon Pileggi, E. David Aja https://rstats.wtf/

-   Happy Git with R for the UseR by Jennifer Bryan, https://happygitwithr.com/

## Additional Resources:: Data/Codes relevant to the Challenge

-   Most information relevant to the challenge is available in the following resources. 
-   [https://localfoodeconomics.com/data/access-the-data/](https://localfoodeconomics.com/data/access-the-data/)
-   [https://allison-bauman.quarto.pub/usda-ams-datametrics/](https://allison-bauman.quarto.pub/usda-ams-datametrics/)
-   [https://github.com/FoodSystemsModeling/DataWarehouse](https://github.com/FoodSystemsModeling/DataWarehouse)
- Most of the examples are already available in R/Quarto.

## Additional Resources:: Useful Packages for Maps/Geospatial Data

Useful packages:

-   `sf`
-   `terra`
-   `tmap`
-   `tidycensus`
-   `leaflet`

## Useful DataViz Books

- [ggplot2: Elegant Graphics for Data Analysis (3e) was written by Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen.](https://ggplot2-book.org/)

- [Drawing beautiful maps programmatically with R, sf and ggplot2 — Part 1: Basics : Oct 25, 2018 • Mel Moreno and Mathieu Basille](https://r-spatial.org/r/2018/10/25/ggplot2-sf.html)

- [R for Data Science (2e) was written by Hadley Wickham, Mine Çetinkaya-Rundel, and Garrett Grolemund.](https://r4ds.hadley.nz/)

- [Modern Data Visualization with R by Robert Kabacoff](https://rkabacoff.github.io/datavis/)

## Useful DataViz Galleries


- [The R Graph Gallery](https://r-graph-gallery.com/)

- [ggplot2 extensions - gallery](https://exts.ggplot2.tidyverse.org/gallery/)

- [R Gallery Book by Kyle W. Brown](https://bookdown.org/content/b298e479-b1ab-49fa-b83d-a57c2b034d49/distributions.html#violin)

- [R CHARTS by R CODER](https://r-charts.com/)

- [The Python Graph Gallery](https://python-graph-gallery.com/)


## Additional Resources: Tools for Interactive Visualizations

- [The R Graph Gallery](https://r-graph-gallery.com/interactive-charts.html)

- [Interactive web-based data visualization with R, plotly, and shiny by Carson Sievert](https://plotly-r.com/)

- [Modern Data Visualization with R by Robert Kabacoff (Chapter 13)](https://rkabacoff.github.io/datavis/)

- [Interactive visualization in R by Rebecca Barter](https://rebeccabarter.com/blog/2017-04-20-interactive)

- [Mastering Shiny by Hadley Wickham](https://mastering-shiny.org/)


## Additional Resources: Quarto


- This revealjs slide deck is created using Quarto, and the .qmd file is available.

- [Quarto Guide](https://quarto.org/docs/guide/)

- [Quarto Tutorial](https://quarto.org/docs/get-started/hello/rstudio.html)



## Acknowledgements

-   Clay Karnis helped me with some of the visualizations.
-   Ivy Mackereth shared her experience from the previous competition.
-   I thank organizers of the webinar for inviting me to share my experience.
-   Thanks everyone for listening, and good luck!

# Any Questions?
